Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-02-20 newpaper/paper.tex
wip

Now, whenever the return value of a $\mbf{case}$-splitting action is a sum of
values (this includes just returning an object value, which is by far the most
common situation), we can use $\join \dlr \case \vx\,\of ...$ to eliminate code
duplication, without creating runtime sum types.

@@

Let us see some operations on streams now. First, $\Pull$ is a ``zippy''
applicative functor, where $\pure$ infinitely repeats a value and $(\!\ap\!)$
applies a stream of functions to a stream of values, pointwise.
\begin{alignat*}{3}
  &\mit{pure} : \vA \to \Pull\,\vA\\
  &\mit{pure} = \Pull\,()\,()\,(\lam \_.\,\return \dlr \Stop)
\end{alignat*}
\begin{alignat*}{3}
  &(\!\ap\!) : \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB \\
  &(\!\ap\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =\\
  & \ind \Pull\,(\vS,\,\vS')\,(\seed,\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
  & \ind \ind \Stop \hspace{1.20em}        \to \return \Stop \\
  & \ind \ind \Skip\,\vs \hspace{0.75em}     \to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \Yield\,\vf\,\vs \to \step'\,\vs' \bind \lam \case \\
  & \ind \ind \ind \Stop \hspace{1.7em} \to \return \Stop \\
  & \ind \ind \ind \Skip\,\vs'\hspace{0.9em}\to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \ind \Yield\,\va\,\vs' \to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs')
\end{alignat*}
In $\mit{pure}$, the state is the unit type, while in $(\!\ap\!)$ we take the
product of the stream states and do synchronous transitions. In both
definitions, the $\IsSOP\,S$ constraint is implicitly dispatched by instance
resolution; this works in the Agda and Haskell versions too. We can derive the
usual $\mit{zip}$ and $\mit{zipWith}$ operations from $(\!\ap\!)$. $\Pull$
is also a monoid with stream appending as the binary operation.
\begin{alignat*}{3}
  & \mit{empty} : \Pull\,\vA\\
  & \mit{empty} = \Pull\,()\,()\,(\lam \_.\,\return \Stop)
\end{alignat*}
\begin{alignat*}{5}
  & \hspace{-4em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
  & \hspace{-4em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left\,\seed)\,\lam \case} \\
  & \hspace{-4em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to \return \dlr \Skip\,(\Right\,\seed')\\
  & \hspace{-4em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
  & \hspace{-4em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
  & \hspace{-4em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
\end{alignat*}
These definitions are standard for streams; note though that compared to the
unstaged definitions in previous literature, the only additional noise is just
the $\Gen$ monad in the transitions. Likewise, we can give standard definitions
for the usual stream functions such as $\msf{filter}$, $\msf{take}$ or
$\msf{drop}$.

@@

How do we generate object code from streams? The $\vS$ state is given as a
finite sums of products, but the sums and the products are on the meta
level, so we cannot directly use $\vS$ in object code. Similarly as in the the
treatment of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$
transition function to a product of functions. However, these functions need to
be \emph{mutually recursive}, since it is possible to transition from any state
to any other state, and each such transition is represented as a function call.
The problem of generating well-typed mutual blocks was addressed by Yallop and
Kiselyov in \cite{TODO}. Unlike ibid., which used control effects and mutable
references in MetaOCaml, we present a solution that does not use side effects
in the metalanguage.

@@

We saw that $\Pull$ is $\Applicative$, but what about having a list-like
$\Monad$ instance as well, with singleton streams for $\return$ and $\concatMap$
for binding? This is not possible. If we have
\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
that means that $\vA \to \Pull\,\vB$ can contain an infinite number of different
machine state types, which can never be represented in a finite amount of object
code.\footnote{Although in two-level type theories that are not used for
staging, the representability of countably infinite products of object types is
sometimes assumed as an axiom; this axiom is called ``cofibrancy of
$\mathbb{N}$'' \cite{TODO}.} Here by ``infinite'' we mean the notion that is
internally available in the meta type-theory. For instance, we can define a
$\Nat_\M \to \Pull\,\vB$ function which for each $\vn : \Nat_\M$ produces a
concatenation of $\vn$ streams. Hence, we shall have the following function
instead:
\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
The idea is the following: if $\USOP$ is closed under dependent sum types, we
can directly define this $\concatMap$, by taking the appropriate dependent sum
of the $\vA \to \USOP$ family of machine states, which we extract from the $\vA
\to \Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent
sums, for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing
and $\fst$ and $\snd$ for projections. We use the following field projection functions:
$\mit{projS} : \Pull\,\vA \to \MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to
\mit{projS}\,\vas$, and $\mit{projStep} : (\vas : \Pull\,\vA) \to
\mit{projS}\,\vas \to \Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.
For now, we assume that the following instance exists:
\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
instance constraint; this is available in Agda and, less generally, in Haskell
\cite{TODO}. The definition of $\concatMap$ is as follows.
\begin{alignat*}{3}
  & \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \\
  & \concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \\
  & \ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(\seed,\,\Nothing) \dlr \lam \case \\
  & \ind \ind (\vs,\,\Nothing) \to \step\,s \bind \lam \case \\
  & \ind \ind \ind \Stop      \hspace{1.4em}   \to \return \Stop\\
  & \ind \ind \ind \Skip\,\vs \hspace{0.9em} \to \return \dlr \Skip\,(\vs,\, \Nothing)\\
  & \ind \ind \ind \Yield\,\va\,\vs          \to \return \dlr \Skip\,(\vs,\, \Just\,(\va,\,\mit{projSeed}\,(\vf\,\va)))\\
  & \ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case \\
  & \ind \ind \ind \Stop       \hspace{1.75em}\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
  & \ind \ind \ind \Skip\,\vs' \hspace{0.94em}\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
  & \ind \ind \ind \Yield\,\vb\,\vs' \to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))\\
\end{alignat*}
Here, $\Nothing$ marks the states where we are in the ``outer'' loop, trying to
run the $\Pull\,\vA$ stream until we get the next value. $\Just$ marks the
states of the ``inner'' loop, where we have a concrete $\va : \vA$ value and
we run the $\vf\,\va$ stream until it ends. In the inner loop, the machine state
type depends on the $\va : \vA$ value, hence the need for $\Sigma$.

@@

Since $\Pull$ is not a monad, it cannot be $\MonadGen$ in our framework,
but we still support some forms of let-insertion. First, we note that $\Pull$ can be
mapped over by functions returning in $\Gen$:
\begin{alignat*}{3}
  &\hspace{-3em}\mathrlap{\mit{mapGen} : \Pull\,\vA \to (\vA \to \Gen\,\vB) \to \Pull\,\vB }\\
  &\hspace{-3em}\mathrlap{\mit{mapGen}\,(\Pull\,\vS\,\seed\,\step)\,\vf = \Pull\,\vS\,\seed \dlr \lam \case}\\
  &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
  &\hspace{-3em} \ind \Skip\,\vs       &&\to \return \dlr \Skip\,\vs\\
  &\hspace{-3em} \ind \Yield\,\va\,\vs &&\to \mdo \{\vb \fro \vf\,\va; \return \dlr \Yield\,\vb\,\vs\}
\end{alignat*}
Now, for some $\vas : \Pull\,(\Lift \vA)$, we may use
\[ \mit{mapGen}\,\vas\,\gen : \Pull\,(\Lift \vA) \]
to let-bind all yielded expressions in the stream. This can be useful when some
stream operations duplicate yielded values. For example, the function with type
$\Pull\,\vA \to \Pull\,(\vA,\,\vA)$ is often used to avoid duplicating whole streams,
and when $\vA$ is an object type we can use let-insertion in the definition to also
avoid duplicating object expressions.

Let us now consider a more ``monadic'' style of stream programming. For this we
need singleton streams:
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
We also write $\mit{forEach}$ for $\concatMap$ with flipped arguments. Now, a
more monadic flavor of let-insertion is given as
\begin{alignat*}{3}
  &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
  &\gen_{\Pull}\,\va = \mit{mapGen}\,(\mit{single}\,\va)\,\gen.
\end{alignat*}
We can also implement case splitting on object values. Recall from Section
\ref{TODO} that for each value type we have a splitting function that returns a
meta-level view on the constructors. Importantly, all such views are $\SOP$-s, and
therefore can be bound by $\forEach$. A small example:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\forEach\,(\mit{countFrom}\,\qt{0}) \dlr \lam \vn. }\\
  & \hspace{-5em}\mathrlap{\forEach\,(\gen_{\Pull}\,\qt{\spl \vn * 2}) \dlr \lam \vm. }\\
  & \hspace{-5em}\mathrlap{\forEach\,(\single\,(\msplit\,\qt{\spl \vm < 20})) \dlr \lam \case }\\
  & \hspace{-5em}\ind \True_\M  &&\to \single\,\qt{\spl \vm + 10}   \\
  & \hspace{-5em}\ind \False_\M &&\to \single\,\qt{\spl \vm + 20}
\end{alignat*}
Here $(\gen_{\Pull}\,\qt{\spl \vn * 2})$ behaves like a let-insertion, and
$(\single\,(\msplit\,\qt{\spl \vm < 20}))$ implements the case splitting.

@@

In the Haskell, there are some limitations. First, in $\concatMap$, the
projection function $\mit{projS}$ cannot be defined, nbecause Haskell is not
dependently typed, and the other field projections are also out of reach. We
only have a weaker ``positive'' recursion principle for existential types. It
might be the case that a strongly typed $\concatMap$ is possible with only weak
existentials, but we attempted this and found that it introduces too much
technical complication.

@@

\section{TODOS}

\todo{Introduce newtypes accordingly}
\todo{allow arbitrary parameters (including ValTy -> ValTy functions) for object ADTs}
\todo{extend as I have more stuff}
\todo{FIGURE OUT: monadic tailrec}
\todo{FIGURE OUT: Traversals}
\todo{DEVELOP: mutual letrec based on computational product types}

%% \section{The Object Language}\label{sec:the-object-language}

%% Describe object language. Simple types + first-order fun + fixpoints + ADT-s.

%% Show Agda formalization basically, types and syntax.

%% Show Agda definitional interpreter. Discuss closure-free evaluation

%% Examples, properties. Maybe big-step reduction? Maybe program equivalence?  Look
%% at Harper PFPL for program equivalence reference.

%% Sketch downstream compilation.
%% \begin{itemize}
%% \item Eta-expand function definitions to arity.
%% \item Turn functions in immediate beta-redexes to let-bindings.
%% \item Lambda-lift non-tail-called functions to top.
%% \item Keep only-tail-called local functions as join points.
%% \end{itemize}

%% Downstream compilation should erase the unit type and implement products as
%% ``flat'' unboxed structures.

%% Discuss object language design choices.
%% \begin{itemize}
%% \item Can be more obviously first-order: e.g.\ no currying, only Val->Val
%%   functions, or ANF functions. This requires more bureaucracy in
%%   metaprogramming.
%% \item Can be more flexible but less obviously first-order: allow function let
%%   bodies, allow function case bodies. Requires more translation. More convenient
%%   to program in. In future work.
%% \end{itemize}

%% \section{The Staged Language}\label{sec:the-staged-language}

%% Don't give horizontal-bar rules.

%% Meta-features:
%% \begin{itemize}
%% \item Pi, Sigma, Unit, Bottom, Universes (indices elided), ADT-s. Ind families probably not needed.
%% \item Agda notation.
%% \item Type classes informally in Haskell notation?
%% \end{itemize}

%% Explain staging ops, staging algorithm. Explain object embedding. Refer to 2LTT
%% paper for correctness.

%% Mention stage annotation inference, but we'll not use it often! To make things
%% explicit.

%% We should include a generous amount of staging examples, and some traced staging.

%% Basic staging examples: identity, map, power function. Boolean short-circuiting.

%% Basic staged classes, Eq, Ord, Show, Monoid, etc. Vectors with computed length.

%% \section{Programming With Monads}\label{sec:programming-with-monads}

%% \subsection{Basic Binding Time Improvements}\label{sec:basic-binding-time-improvements}

%% Notion.
%% For functions.
%% For products.
%% Note that for products there is a computation duplication problem.
%% Mention let-insertion.

%% \subsection{The Code Generation Monad}\label{sec:the-code-generation-monad}

%% Is fundamental to this work.

%% Definition. Explanation. Define Functor, Applicative, Monad instances.
%% Define runGen, runGen1, runGenN etc. Define let-insertion.

%% Basic examples.

%% Explanation in terms of Church-coding.

%% Gen Vector mapping or folding example from 2LTT repo. Without Gen, it's a
%% big pain to get ideal linear-size map code in N. With Gen it's easy-peasy.

%% \subsection{Monad Transformers}\label{sec:monad-transformers}

%% Identity, ReaderT, StateT, MaybeT, ExceptT

%% Strict versions of put, modify, local, runner functions.

%% MTL-style overloading. MonadReader, MonadState, MonadError, MonadGen, liftGen

%% Binding time improvement for monads. Up/down overloading.

%% \subsection{Case Splitting in MonadGen}\label{sec:case-splitting-in-monadgen}

%% It works for any object ADT.

%% Introduce sums-of-values (SOV).

%% Introduce syntax sugar for object case splits in MonadGen blocks.

%% Basic examples.

%% \subsection{Joining Control Flow}\label{sec:rejoining-control-flow}

%% The problem of join points. Motivation in terms of MaybeT Gen.

%% We can let-bind computations by tripping through up/down of monads,
%% but that introduces unbox/rebox cost for Maybe \& Either.

%% Works OK for Reader/State though.

%% \begin{verbatim}
%% f : Bool -> MaybeT0 Gen ^Int
%% f = \b. ~(down do
%%   x <- case <b> of
%%     True  -> gen <10>
%%     False -> gen <20>
%%   y <- gen <~x + 10>
%%   z <- gen <~x * 10>
%%   pure <~x + ~y + ~z>)
%% \end{verbatim}

%% Join points using SOV-s. Description of SOV machinery. MonadJoin.
%% instance MonadJoin Gen. MonadJoin MaybeT.

%% \subsection{Monadic Tail Recursion}\label{sec:monadic-tail-recursion}

%% \texttt{replicateM\_}, \texttt{find} functions working in ``any'' monad.
%% Non-tail recursion is not possible to fuse, return addresses are dynamic.
%% Tail recursion can return to statically known points.

%% \subsection{Discussion}

%% Limitations. We can't put non-value actions into structures, can't accumulate them.

%% Guarantees/properties. Closure-freedom. No administrative redexes. No boxing in join points.
%% Only boxing in monadic code comes from ``down''. No code blowup.

%% \section{Fusion}\label{sec:push-pull-fusion}

%% Motivation.

%% \subsection{Push streams}\label{sec:push-streams}

%% Definition.

%% \subsection{Pull streams}\label{sec:pull-streams}

%% Pull streams are definitely not monadic. \texttt{Nat -> Stream a} contains
%% ``infinite'' amount of code internally.

%% Pull streams are also not selective functors.

%% We do have monadic binding for lifted types, plus ``case'' matching
%% on arbitrary object values! This is practically almost as good as
%% having unlimited monadic bind! We need to assume a ``non-dependency axiom'',
%% that any $f : \Lift A \to \msf{ValTy}$ must be constant.

%% \todo{Validate non-dependency in the psh model!}
%% \todo{Add meta-identity type to the psh model}

\interlinepenalty=10000
\bibliography{references}

>>>>

Now, whenever the return value of a $\mbf{case}$-splitting action is a sum of
products of values (this includes just returning an object value, which is by
far the most common situation), we can use $\join \dlr \case \vx\,\of ...$ to
eliminate code duplication, without creating runtime sum types.

@@

Let us see some operations on streams now. First, $\Pull$ is a ``zippy''
applicative functor, where $\mit{pure}$ infinitely repeats a value and $({\ap})$
applies a stream of functions to a stream of values, pointwise.
\begin{alignat*}{3}
  &\mit{pure} : \vA \to \Pull\,\vA\\
  &\mit{pure} = \Pull\,()\,()\,(\lam \_.\,\return \dlr \Stop)
\end{alignat*}
\begin{alignat*}{3}
  &(\!\ap\!) : \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB \\
  &(\!\ap\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =\\
  & \ind \Pull\,(\vS,\,\vS')\,(\seed,\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
  & \ind \ind \Stop \hspace{1.20em}        \to \return \Stop \\
  & \ind \ind \Skip\,\vs \hspace{0.75em}     \to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \Yield\,\vf\,\vs \to \step'\,\vs' \bind \lam \case \\
  & \ind \ind \ind \Stop \hspace{1.7em} \to \return \Stop \\
  & \ind \ind \ind \Skip\,\vs'\hspace{0.9em}\to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \ind \Yield\,\va\,\vs' \to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs')
\end{alignat*}
In $\mit{pure}$, the state is the unit type, while in $(\!\ap\!)$ we take the
product of the stream states and do synchronous transitions. In both
definitions, the $\IsSOP\,\vS$ constraint is implicitly dispatched by instance
resolution; this works in the Agda and Haskell versions too. We can derive the
usual $\mit{zip}$ and $\mit{zipWith}$ operations from $(\!\ap\!)$. $\Pull$
is also a monoid with stream appending as the binary operation.
\begin{alignat*}{3}
  & \mit{empty} : \Pull\,\vA\\
  & \mit{empty} = \Pull\,()\,()\,(\lam \_.\,\return \Stop)
\end{alignat*}
\begin{alignat*}{5}
  & \hspace{-4em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
  & \hspace{-4em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left\,\seed)\,\lam \case} \\
  & \hspace{-4em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to \return \dlr \Skip\,(\Right\,\seed')\\
  & \hspace{-4em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
  & \hspace{-4em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
  & \hspace{-4em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
\end{alignat*}
These definitions are standard for streams; note though that compared to the
unstaged definitions in previous literature, the only additional noise is just
the $\Gen$ monad in the transitions. Likewise, we can give standard definitions
for usual stream functions such as $\msf{filter}$, $\msf{take}$ or
$\msf{drop}$.

@@

How do we generate object code from streams? The $\vS$ state is given as a
finite sums of products, but the sums and the products are on the meta
level, so we cannot directly use $\vS$ in object code. Similarly as in the the
treatment of join points, we tabulate the $\vS \to \Gen\,\Step\,\vS\,\vA$
transition function to a product of functions. However, these functions need to
be \emph{mutually recursive}, since it is possible to transition from any state
to any other state, and each such transition is represented as a function call.
This problem of generating well-typed mutual blocks was addressed by Yallop and
Kiselyov in \cite{TODO}. Unlike ibid., which used control effects and mutable
references in MetaOCaml, we present a solution that does not use side effects
in the metalanguage.

@@

We saw that $\Pull$ is $\Applicative$, but what about having a list-like
$\Monad$ instance as well, with singleton streams for $\return$ and $\concatMap$
for binding? This is not possible. Aiming at
\[ \concatMap : (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
the $\vA \to \Pull\,\vB$ function can contain an infinite number of different
machine state types, which cannot be represented in a finite amount of object
code.\footnote{Although in two-level type theories that are not used for
staging, the representability of countably infinite products of object types is
sometimes assumed as an axiom; this axiom is called ``cofibrancy of
$\mathbb{N}$'' \cite{TODO}.} Here by ``infinite'' we mean the notion that is
internally available in the meta type theory. For instance, we can define a
$\Nat_\M \to \Pull\,\vB$ function which for each $\vn : \Nat_\M$ produces a
concatenation of $\vn$ streams. Hence, we shall have the following function
instead:
\[ \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB, \]
The idea is the following: if $\USOP$ is closed under dependent sum types, we
can directly define this $\concatMap$, by taking the appropriate dependent sum
of the $\vA \to \USOP$ family of machine states, which we extract from the $\vA
\to \Pull\,\vB$ function. Let us write $\Sigma\,\vA\,\vB : \MTy$ for dependent
sums, for $\vA : \MTy$ and $\vB : \vA \to \MTy$, and reuse $({,})$ for pairing.
We also use the following field projection functions: $\mit{projS} : \Pull\,\vA
\to \MTy$, $\mit{projSeed} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas$, and
$\mit{projStep} : (\vas : \Pull\,\vA) \to \mit{projS}\,\vas \to
\Gen\,(\Step\,(\mit{projS}\,\vas)\,\vA)$.  For now, we assume that the following
instance exists:
\[ \instance (\IsSOP\,\vA,\,\{\va : \vA\} \to \IsSOP\,(\vB\,\va)) \RA \IsSOP\,(\Sigma\,\vA\,\vB) \]
Above, $\{\va : \vA\} \to \IsSOP\,(\vB\,\va)$ is a universally quantified
instance constraint; this is available in Agda and, less generally, in Haskell
\cite{TODO}. The definition of $\concatMap$ is as follows.
\begin{alignat*}{3}
  & \concatMap : \IsSOP\,\vA \RA (\vA \to \Pull\,\vB) \to \Pull\,\vA \to \Pull\,\vB \\
  & \concatMap\,\{\vA\}\,\{\vB\}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \\
  & \ind \Pull\,(\vS,\,\Maybe\,(\Sigma\,\vA\,(\mit{projS} \circ \vf)))\,(\seed,\,\Nothing) \dlr \lam \case \\
  & \ind \ind (\vs,\,\Nothing) \to \step\,s \bind \lam \case \\
  & \ind \ind \ind \Stop      \hspace{1.4em}   \to \return \Stop\\
  & \ind \ind \ind \Skip\,\vs \hspace{0.9em} \to \return \dlr \Skip\,(\vs,\, \Nothing)\\
  & \ind \ind \ind \Yield\,\va\,\vs          \to \return \dlr \Skip\,(\vs,\, \Just\,(\va,\,\mit{projSeed}\,(\vf\,\va)))\\
  & \ind \ind (\vs,\,\Just\,(\va,\,\vs')) \to \mit{projStep}\,(\vf\,\va)\,\vs' \bind \lam \case \\
  & \ind \ind \ind \Stop       \hspace{1.75em}\to \return \dlr \Skip\,(\vs,\,\Nothing) \\
  & \ind \ind \ind \Skip\,\vs' \hspace{0.94em}\to \return \dlr \Skip\,(\vs,\,\Just\,(\va,\,\vs'))\\
  & \ind \ind \ind \Yield\,\vb\,\vs' \to \return \dlr \Yield\,\vb\,(\vs,\, \Just\,(\va,\,\vs'))\\
\end{alignat*}
Here, $\Nothing$ marks the states where we are in the ``outer'' loop, trying to
run the $\Pull\,\vA$ stream until we get the next value. $\Just$ marks the
states of the ``inner'' loop, where we have a concrete $\va : \vA$ value and
we run the $(\vf\,\va)$ stream until it stops. In the inner loop, the machine state
type depends on the $\va : \vA$ value, hence the need for $\Sigma$.

@@

Since $\Pull$ is not a monad, it cannot be $\MonadGen$ in our framework,
but we still support some forms of let-insertion. First, we note that $\Pull$ can be
mapped over by functions returning in $\Gen$:
\begin{alignat*}{3}
  &\hspace{-3em}\mathrlap{\mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
  &\hspace{-3em}\mathrlap{\mit{mapGen}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \Pull\,\vS\,\seed \dlr \lam \case}\\
  &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
  &\hspace{-3em} \ind \Skip\,\vs       &&\to \return \dlr \Skip\,\vs\\
  &\hspace{-3em} \ind \Yield\,\va\,\vs &&\to \mdo \{\vb \fro \vf\,\va; \return \dlr \Yield\,\vb\,\vs\}
\end{alignat*}
Now, we may use $\mit{mapGen}\,\gen : \Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA)$
to let-bind all yielded expressions in a stream. This can be useful in stream
operations that would otherwise duplicate yielded expressions. For example, the
function with type $\Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA,\,\Lift \vA)$
should be defined with let-insertion.

Let us now consider a more ``monadic'' style of stream programming. For this we
need singleton streams:
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
We also write $\mit{forEach}$ for $\concatMap$ with flipped arguments. Now, a
more monadic flavor of let-insertion is given as
\begin{alignat*}{3}
  &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
  &\gen_{\Pull}\,\va = \mit{mapGen}\,\gen\,(\mit{single}\,\va).
\end{alignat*}
We can also implement case splitting on object values. Recall from Section
\ref{TODO} that for each value type we have a splitting function that returns a
meta-level view on the constructors. Importantly, all such views are $\SOP$-s, and
therefore can be bound by $\forEach$. A small example:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\forEach\,(\mit{countFrom}\,\qt{0}) \dlr \lam \vn. }\\
  & \hspace{-5em}\mathrlap{\forEach\,(\gen_{\Pull}\,\qt{\spl \vn * 2}) \dlr \lam \vm. }\\
  & \hspace{-5em}\mathrlap{\forEach\,(\single\,(\msplit\,\qt{\spl \vm < 20})) \dlr \lam \case }\\
  & \hspace{-5em}\ind \True_\M  &&\to \single\,\qt{\spl \vm + 10}   \\
  & \hspace{-5em}\ind \False_\M &&\to \single\,\qt{\spl \vm + 20}
\end{alignat*}
Here $(\gen_{\Pull}\,\qt{\spl \vn * 2})$ behaves like a let-insertion, and
$(\single\,(\msplit\,\qt{\spl \vm < 20}))$ implements the case splitting.

@@

In Haskell there are some limitations. First, in $\concatMap$, the
projection function $\mit{projS}$ cannot be defined, because Haskell is not
dependently typed, and the other field projections are also out of reach. We
only have a weaker ``positive'' recursion principle for existential types. It
might be the case that a strongly typed $\concatMap$ is possible with only weak
existentials, but we attempted this and found that it introduces too much
technical complication.

@@

%% \interlinepenalty=10000
\bibliography{references}
